---
title: "Differential Expression testing"
output: html_document
date: "2024-03-22"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(SCpubr)
```

## Differential gene expression between samples and across conditions

In some cases we will have access to datasets of many single cell samples and we want to compare the expression between these. 
While we can use Integration analysis and Clustering to deal with batch effects, we should not use `FindMarkers` to compare the expression between samples.
Because `FindMarkers` is designed to compare the expression between clusters, we do not model effects introduced by each sample in the test. 

To date one of the simplest and effective ways to compare gene expression is to pseudobulk the data.
This means that we aggregate counts from all cells from a single sample for each cell type and then compare the expression between samples.

The newest version of Seurat allows us to do this with the `AggregateExpression` function.

```{r cars}
sample <- readRDS("../../data/GSE205771/GSM6226122_oligodendroglioma_seurat_object.rds")
sample$orig.ident %>% unique()

paste0(sample$orig.ident, "_", sample$grade) %>% unique()

do_DimPlot(sample, reduction = "umap")
```

## Aggergate Expression and run test

The code below shows how to run this analysis for a given cell Type in the dataset.
Note, that in the group.by arguemtn we that we are grouping by:
- grade -> Tumor grade of the patient
- orig.ident -> The original sample ID of the patient
- New_NMF_labelling -> The cell type of the cell

After aggregation we can run FindMarkers using "DESeq2" as a test since after Pseudobulking 
are data can be treated as Bulk RNA-seq data.

```{r pressure, echo=FALSE}
pseudobulked_seurat <- AggregateExpression(sample, assays = "RNA", return.seurat = T, group.by = c("grade", "orig.ident", "New_NMF_labelling"))
pseudobulked_seurat$celltype.grade <- paste(pseudobulked_seurat$New_NMF_labelling, pseudobulked_seurat$grade, 
                                            sep = "_")
tail(Cells(pseudobulked_seurat))

Idents(pseudobulked_seurat) <- "celltype.grade"

cell_type <- "Cycling-like"
bulk_celltype_de <- FindMarkers(object = pseudobulked_seurat, 
                         ident.1 = glue::glue("{cell_type}_g2"), 
                         ident.2 = glue::glue("{cell_type}_g3"),
                         test.use = "DESeq2")
head(bulk_celltype_de, n = 15)
```

For comparison we also do differential expression at a single cell level: 

```{r}
sample$celltype_grade <- paste(sample$New_NMF_labelling, sample$grade, sep = "_")
Idents(sample) <- "celltype_grade"
celltype_de <- FindMarkers(sample, ident.1 = glue::glue("{cell_type}_g2"),
                                    ident.2 = glue::glue("{cell_type}_g3"),
                                    verbose = FALSE)
head(celltype.de, n = 10)
```

Now we can compare the results:


> ⌨🔥 Exercise:  compare the results between the groups hint (Look at p-value and LogFoldChanges between groups)
> Run this analysis for many cell types (or design a loop to run it for each cell type ?)
>k
> Organize and look at your results. 
> How would you visuliaze your results?


As a propsed way to identify consistently changing genes, we can look at Genes
that are differentially expressed in both the single cell and pseudobulk level.

```{r}
# compare the DE P-values between the single-cell level and the pseudobulk level results
names(bulk_celltype_de) <- paste0(names(bulk_celltype_de), ".bulk")
bulk_celltype_de$gene <- rownames(bulk_celltype_de)

names(celltype_de) <- paste0(names(celltype_de), ".sc")
celltype_de$gene <- rownames(celltype_de)

merge_dat <- merge(celltype_de, bulk_celltype_de, by = "gene")
merge_dat <- merge_dat[order(merge_dat$p_val.bulk), ]

# Number of genes that are marginally significant in both; marginally significant only in bulk; and marginally significant only in single-cell
common <- merge_dat$gene[which(merge_dat$p_val.bulk < 0.05 & 
                                merge_dat$p_val.sc < 0.05)]
only_sc <- merge_dat$gene[which(merge_dat$p_val.bulk > 0.05 & 
                                  merge_dat$p_val.sc < 0.05)]
only_bulk <- merge_dat$gene[which(merge_dat$p_val.bulk < 0.05 & 
                                    merge_dat$p_val.sc > 0.05)]
print(paste0('# Common: ',length(common)))
```

> ⌨🔥 Exercise:  Visualize the results of the comparison between the single cell and pseudobulk level results
> Which genes are conserved in DE expression which are not?
> Visualize top differentially expressed genes in the original data (VlnPlot or UMAP featureplots)? What does this tell us?

```{r}
do_FeaturePlot(sample, features = "CD36")
sample$celltype.grade <- paste(sample$New_NMF_labelling, sample$grade, 
                                            sep = "_")
Idents(sample) <- "celltype.grade"
celltype_de <- FindMarkers(sample, ident.1 = "Gradient_2", ident.2 = "Gradient_3", verbose = FALSE)

VlnPlot(sample, features = "KCND2", idents = c("Gradient_2", "Gradient_3"), group.by = "grade") 
```


Lastly, there are a bunch of methods and packages for differential gene expression analysis.
In Seurat alone there are 10 different methods to chose from. https://satijalab.org/seurat/reference/findmarkers
Which test method you choose will depend on data structure and biological question. 
The advantage of Pseudobulk analysis is that we avoid inflated False Positives due to the high number of cells. 


Further reading: https://www.nature.com/articles/s41467-021-25960-2
https://www.nature.com/articles/s41467-023-37126-3#Sec15